---
title: "Alpha design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE, purl=FALSE}
# formatting tables for html output
knitr::opts_chunk$set(message = FALSE, warning = FALSE, purl = TRUE)
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
```

```{r}
# packages
pacman::p_load(readr, tidyverse, forcats, # data import and handling
               lme4, lmerTest,            # linear mixed model 
               emmeans, multcomp,         # mean comparisons
               ggplot2, desplot)          # plots
```

# Data

This example is taken from Chapter *"3.8 Analysis of an $\alpha$-design"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [John and Williams (1995)](https://onlinelibrary.wiley.com/doi/abs/10.1002/bimj.4710380703){target="_blank"} from a yield (t/ha) trial laid out as an alpha design. The trial had 24 genotypes (`gen`), 3 complete replicates (`rep`) and 6 incomplete blocks (`inc.block`) within each replicate. The block size was 4.

## Import

```{r}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/John%26Williams1995.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting
Before anything, the columns `plot`, `rep`, `inc.block` and `gen` should be encoded as factors, since R by default encoded them as character.

```{r}
dat <- dat %>% 
  mutate_at(vars(plot:gen), as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `column` of each plot in the trial. 

```{r, fig.height = 3, fig.align = "center"}
desplot(data = dat,
        form = gen ~ col + row | rep,          # fill color per genotype, headers per replicate
        text = gen, cex = 0.7, shorten = "no", # show genotype names per plot
        out1 = rep,                            # lines between complete blocks/replicates
        out2 = inc.block,                      # lines between incomplete blocks
        main = "Field layout", show.key = F)   # formatting
```

An $\alpha$-design is a design with incomplete blocks, where the blocks can be grouped into complete replicates. Such designs are termed “resolvable”. The model must have an effect for complete replicates, and effects for incomplete blocks must be nested within replicates.

We could also have a look at the arithmetic means and standard deviations for yield per genotype (`gen`) or incomplete block (`inc.block`). Notice that the way the factor variable `inc.block` is defined, it only has 6 levels (`r levels(dat$inc.block)`). However, as can be clearly seen on the field layout above, there are 18 incomplete blocks, *i.e.* 6 per replicate. Thus, an actual incomplete block here is identified by the information stored in `rep` *and* `inc.block`. Alternatively, one could have coded the `inc.block` variable as a factor with 18 levels (B1-B18), but this was not done here.

<div class = "row"> <div class = "col-md-6">
```{r}
dat %>% 
  group_by(gen) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(desc(mean)) %>% # sort
  print(n=Inf) # print full table
```
</div> <div class = "col-md-6">
```{r}
dat %>% 
  group_by(rep, inc.block) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield))  %>% 
  arrange(desc(mean)) %>% # sort
  print(n=Inf) # print full table
```
</div> </div>

We can also create a plot to get a better feeling for the data.

```{r, fig.align = "center"}
plotdata <- dat %>% 
  group_by(gen) %>% 
  mutate(mean_yield = mean(yield)) %>% # add column with mean yield per gen
  ungroup() %>% 
  mutate(gen = fct_reorder(.f = gen, .x = mean_yield)) # sort factor variable by mean yield

ggplot(data = plotdata, 
       aes(x = gen)) +
  geom_point(aes(y = yield, shape = rep)) +  # scatter plot observed
  geom_point(aes(y = mean_yield), color = "cornflowerblue") + # scatter plot mean
  ylim(0, NA) +   # force y-axis to start at 0
  labs(caption = "Blue dots represent arithmetic mean per genotype") +
  theme_classic() + # clearer plot format 
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) # rotate x-axis labels
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable and (fixed) `gen` and `block` effects. There also needs to be term for the 18 incomplete blocks (*i.e.* `rep:inc.block`) in the model, but it can be taken either as a fixed or a random effect. Since our goal is to compare genotypes, we will determine which of the two models we prefer by comparing the average standard error of a difference (s.e.d.) for the comparisons between adjusted genotype means - the lower the s.e.d. the better.

<div class = "row"> <div class = "col-md-6">
```{r}
# blocks as fixed (linear model)
mod.fb <- lm(yield ~ gen + rep +
               rep:inc.block,
             data = dat)

mod.fb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pull("SE") %>% # extract s.e.d. column
  mean() # get arithmetic mean
```
</div> <div class = "col-md-6">
```{r}
# blocks as random (linear mixed model)
mod.rb <- lmer(yield ~ gen + rep +
                 (1 | rep:inc.block),
               data = dat)

mod.rb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pull("SE") %>% # extract s.e.d. column
  mean() # get arithmetic mean
```
</div> </div>

As a result, we find that the model with random block effects has the smaller s.e.d. and is therefore more precise in terms of comparing genotypes.

## Variance component estimates

We can extract the variance component estimates for our mixed model as follows:

```{r}
mod.rb %>% 
  VarCorr() %>% 
  as.data.frame() %>% 
  dplyr::select(grp, vcov)
```

## ANOVA

Thus, we can conduct an ANOVA for this model. As can be seen, the F-test of the ANOVA (using Kenward-Roger’s method for denominator degrees-of-freedom and F-statistic) finds the `gen` effects to be statistically significant (p`r anova(mod.rb, ddf="Kenward-Roger")["gen", "Pr(>F)"] %>% format.pval(pv = ., eps = .001, digits = 2)`). 

```{r anova}
mod.rb %>% anova(ddf="Kenward-Roger")
```

## Mean comparisons

```{r}
mean_comparisons <- mod.rb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("emmeans") %>%
  cld(details = TRUE, Letters = letters) # add letter display

mean_comparisons$emmeans # adjusted genotype means
```

# Present results

## Mean comparisons

For this example we can create a plot that displays both the raw data and the results, *i.e.* the comparisons of the adjusted means that are based on the linear model.

```{r}
ggplot() +
  # black dots representing the raw data
  geom_point(
    data = plotdata,
    aes(y = yield, x = gen)
  ) +
  # red dots representing the adjusted means
  geom_point(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = gen),
    color = "red",
    position = position_nudge(x = 0.1)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = mean_comparisons$emmeans,
    aes(ymin = lower.CL, ymax = upper.CL, x = gen),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.1)
  ) +
  # red letters 
  geom_text(
    data = mean_comparisons$emmeans,
    aes(y = lower.CL, x = gen, label = .group),
    color = "red",
    angle = 90,
    hjust = 1,
    position = position_nudge(y = - 0.1)
  ) + 
  ylim(0, NA) + # force y-axis to start at 0
  ylab("Yield in t/ha") + # label y-axis
  xlab("Genotype") +      # label x-axis
  labs(caption = "Black dots represent raw data
       Red dots and error bars represent adjusted mean with 95% confidence limits per genotype
       Means followed by a common letter are not significantly different according to the Tukey-test") +
  theme_classic() + # clearer plot format 
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) # rotate x-axis
```

**R-Code and exercise solutions**

Please [click here](https://github.com/SchmidtPaul/DSFAIR/tree/master/Rpurl){target="_blank"} to find a folder with `.R` files. Each file contains

 * the entire R-code of each example combined, including
 * solutions to the respective exercise(s).