---
title: "Augmented design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(readr, tidyverse) # data import and handling
```

# Data

This example is taken from Chapter *"3.7 Analysis of a non-resolvable augmented design"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Peterson (199)](https://www.taylorfrancis.com/books/9780429078491){target="_blank"} from a yield trial laid out as an augmented design. The genotypes (`gen`) include 3 standards (`st`, `ci`, `wa`) and 30 new cultivars of interest. The trial was laid out in 6 blocks (`block`). The 3 standards are tested in each block, while each entry is tested in only one of the blocks. Therefore, the blocks are *"incomplete blocks"*.


