---
title: "Augmented design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(janitor, readr, tidyverse, # data import and handling
               lme4,                      # linear mixed model 
               emmeans, multcomp,         # mean comparisons
               ggplot2, desplot)          # plots 
```

# Data

This example is taken from Chapter *"3.7 Analysis of a non-resolvable augmented design"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Peterson (199)](https://www.taylorfrancis.com/books/9780429078491){target="_blank"} from a yield trial laid out as an augmented design. The genotypes (`gen`) include 3 standards (`st`, `ci`, `wa`) and 30 new cultivars of interest. The trial was laid out in 6 blocks (`block`). The 3 standards are tested in each block, while each entry is tested in only one of the blocks. Therefore, the blocks are *"incomplete blocks"*.

## Import

```{r, message=FALSE, warning=FALSE}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Pattersen1994.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting
Before anything, the columns `gen` and `block` should be encoded as factors, since R by default encoded them as character.

```{r}
dat <- dat %>% 
  mutate_at(vars(gen, block), as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `column` of each plot in the trial. 

Notice that for this example it makes sense to put in some extra effort to explicitly assign colors to the genotypes in a way that differntiates the standards (`st`, `ci`, `wa`) from the other genotypes.

```{r, fig.height = 5, fig.align = "center"}
# create a color-generating function
get_shades_of_blue <- colorRampPalette(colors=c("skyblue", "royalblue4"))
blue_30  <- get_shades_of_blue(30) # 30 shades of blue
orange_3 <- c("darkorange", "darkorange2", "darkorange3") # 3 shades of orange
blue_30_and_orange_3 <- c(blue_30, orange_3) # combine into single vector                  

desplot(data = dat,
        form = gen ~ col + row, # fill color per genotype, headers per replicate
        col.regions = blue_30_and_orange_3,  # custom colors
        text = gen, cex = 1, shorten = "no", # show genotype names per plot
        out1 = block,                        # lines between complete blocks/replicates
        main = "Field layout", show.key = F) # formatting
```

We could also have a look at the arithmetic means and standard deviations for yield per genotype (gen) or block (inc.block). It must be clear, however, that except for the standards (`st`, `ci`, `wa`), there is always only a single value per genotype, so that the mean simply is that value and a standard error cannot be calculated at all.

<div class = "row"> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(gen) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(std.dev, desc(mean)) %>% # sort
  print(n=Inf) # print full table
```
</div> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(block) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield))  %>% 
  arrange(desc(mean)) %>% # sort
  print(n=Inf) # print full table
```
</div> </div>

```{r}
ggplot(data = dat, 
       aes(y = yield,
           x = gen,
           color = gen,
           shape = block)) +
  geom_point(size = 2) + # scatter plot with larger dots
  ylim(0, NA) + # force y-axis to start at 0
  scale_color_manual(values = blue_30_and_orange_3) + # custom colors
  guides(color = FALSE) + # turn off legend for colors
  theme_classic() + # clearer plot format
  theme(legend.position = "top") # legend on top
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable and `gen` as fixed effects, since our goal is to compare them to each other. Since the trial was laid out in blocks, we also need `block` effects in the model, but these can be taken either as a fixed or as random effects. Since our goal is to compare genotypes, we will determine which of the two models we prefer by comparing the average standard error of a difference (s.e.d.) for the comparisons between adjusted genotype means - the lower the s.e.d. the better. 

<div class = "row"> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
# blocks as fixed (linear model)
mod.fb <- lm(yield ~ gen + block,
             data = dat)

mod.fb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pluck("SE") %>% # extract s.e.d. column
  mean # get arithmetic mean
```
</div> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
# blocks as random (linear mixed model)
mod.rb <- lmer(yield ~ gen + (1 | block),
               data = dat)

mod.rb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pluck("SE") %>% # extract s.e.d. column
  mean # get arithmetic mean
```
</div> </div>

As a result, we find that the model with fixed block effects has the slightly smaller s.e.d. and is therefore more precise in terms of comparing genotypes.

## ANOVA

Thus, we can conduct an ANOVA for this model.

```{r}
mod.fb %>% anova()
```

As can be seen, the F-test finds the `gen` effects to be statistically significant (p = `r anova(mod.fb)["gen", "Pr(>F)"] %>% round(3)` < 0.05). 

## Mean comparisons

It can be seen that while some genotypes have a higher yield than others, no differences are found to be statistically significant here. Accordingly, notice that *e.g.* for `gen 11`, which is the genotype with the highest adjusted yield mean (=3055), its lower confidence limit (=1587) includes `gen 12`, which is the genotype with the lowest adjusted yield mean (=1632). 

```{r}
mean_comparisons <- mod.fb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("emmeans") %>%
  cld(details = TRUE, Letters = letters) # add letter display

mean_comparisons$emmeans # adjusted genotype means
```

# Present results

## Mean comparisons

For this example we can create a plot that displays both the raw data and the results, *i.e.* the comparisons of the adjusted means that are based on the linear model.

```{r}
ggplot() +
  # blue/orange dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = gen, color = gen)
  ) +
  scale_color_manual(values = blue_30_and_orange_3) + # custom colors
  guides(color = FALSE) + # turn off legend for colors
  # red dots representing the adjusted means
  geom_point(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = gen),
    color = "red",
    position = position_nudge(x = 0.2)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = mean_comparisons$emmeans,
    aes(ymin = lower.CL, ymax = upper.CL, x = gen),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.2)
  ) +
  # red letters 
  geom_text(
    data = mean_comparisons$emmeans,
    aes(y = lower.CL, x = gen, label = .group),
    color = "red",
    angle = 90,
    hjust = 1,
    position = position_nudge(y = - 25)
  ) + 
  ylim(0, NA) + # force y-axis to start at 0
  ylab("Yield in t/ha") + # label y-axis
  xlab("Genotype") +      # label x-axis
  labs(caption = "Blue/orange dots represent raw data
       Red dots and error bars represent adjusted mean with 95% confidence limits per genotype
       Means followed by a common letter are not significantly different according to the Tukey-test") +
  theme_classic() # clearer plot format 
```

# Exercises

## Exercise 1

TThis example is taken from Chapter *"3.9 Analysis of a resolvable design with checks"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data from an augmented design that was laid out for 90 entries and 6 checks. The block size was 10. Incomplete blocks were formed according to a 10 x 10 lattice design, in which incomplete blocks can be grouped into complete replicates. Thus, this is a resolvable design. Checks are coded as 1001 to 1006, while the 90 entries are coded as 2 to 100 (note that there are no entries with labels 11, 21, 31, 41, 51, 61, 71, 81 and 91). Some of the checks have extra replication. We here consider the trait “yield”.

* Explore
  + Draw a plot with yield per genotype
* Analyze
  + Set up two models: One with blocks as fixed and one with blocks as random. Compare their average s.e.d. between genotype means. Choose the model with the smaller value.
  + Compute an ANOVA for the chose model.
  + Perform multiple (mean) comparisons based on the chosen model.

```{r, message=FALSE, warning=FALSE}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/PiephoAugmentedLattice.csv"
ex1dat <- read_csv(dataURL)

desplot(data = ex1dat,
        form = block ~ col + row | rep, # fill color per block, headers per replicate
        text = geno, cex = 0.75, shorten = "no", # show genotype names per plot
        col  = genoCheck, # different color for check genotypes
        out1 = rep,   out1.gpar = list(col = "black"), # lines between reps
        out2 = block, out2.gpar = list(col = "darkgrey"), # lines between blocks
        main = "Field layout", show.key = F) # formatting
```