---
title: "Completely randomized design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(readr, tidyverse, forcats,  # data import and handling
               emmeans, multcomp,          # mean comparisons
               ggplot2, desplot)           # plots
```

# Data

This example is taken from *"Example 4.3"* of the course material ["Quantitative Methods in Biosciences (3402-420)"](https://www.uni-hohenheim.de/en/module-catalogue/modul/quantitative-methods-in-biosciences-3402-420){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Mead et al. (1993, p.52)](https://www.routledge.com/Statistical-Methods-in-Agriculture-and-Experimental-Biology/Mead/p/book/9781584881872){target="_blank"} from a yield trial with melons. The trial had 4 melon varieties (`variety`). Each variety was tested on six field plots. The allocation of treatments (varieties) to experimental units (plots) was completely at random. Thus, the experiment was laid out as a completely randomized design (CRD).

## Import

```{r, message=FALSE, warning=FALSE}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Mead1993.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting
Before anything, the column `variety` should be encoded as a factor, since R by default encoded it as a character variable.

```{r}
dat <- dat %>% 
  mutate_at(vars(variety), as.factor)
```


