---
title: "Completely randomized design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE, purl=FALSE}
# formatting tables for html output
knitr::opts_chunk$set(message = FALSE, warning = FALSE, purl = TRUE)
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
```

```{r}
# packages
pacman::p_load(tidyverse,        # data import and handling
               conflicted,       # handling function conflicts
               emmeans, multcomp, multcompView, # mean comparisons
               ggplot2, desplot) # plots
```

# Data

This example is taken from *"Example 4.3"* of the course material ["Quantitative Methods in Biosciences (3402-420)"](https://www.uni-hohenheim.de/en/module-catalogue/modul/quantitative-methods-in-biosciences-3402-420){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Mead et al. (1993, p.52)](https://www.routledge.com/Statistical-Methods-in-Agriculture-and-Experimental-Biology/Mead/p/book/9781584881872){target="_blank"} from a yield trial with melons. The trial had 4 melon varieties (`variety`). Each variety was tested on six field plots. The allocation of treatments (varieties) to experimental units (plots) was completely at random. Thus, the experiment was laid out as a completely randomized design (CRD).

## Import

```{r}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Mead1993.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting
Before anything, the column `variety` should be encoded as a factor, since R by default encoded it as a character variable.

```{r}
dat <- dat %>% 
  mutate_at(vars(variety), as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `column` of each plot in the trial. 

```{r, fig.height = 3, fig.align = "center"}
desplot(data = dat, flip = TRUE,
        form = variety ~ col + row,              # fill color per genotype
        text = variety, cex = 1, shorten = "no", # show genotype names per plot
        main = "Field layout", show.key = F)     # formatting
```

We could also have a look at the arithmetic means and standard deviations per `variety`:

```{r}
dat %>% 
  group_by(variety) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield))
```

We can also create a plot to get a better feeling for the data.

```{r, fig.height = 3, fig.width = 3, fig.align = "center"}
ggplot(data = dat,
       aes(y = yield, x = variety)) +
  geom_point() +  # scatter plot
  ylim(0, NA) +   # force y-axis to start at 0
  theme_classic() # clearer plot format 
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable and (fixed) `variety` effects.

```{r}
mod <- lm(yield ~ variety, data = dat)
```

## ANOVA

Thus, we can conduct an ANOVA for this model. As can be seen, the F-test of the ANOVA finds the `variety` effects to be statistically significant (p`r anova(mod)["variety", "Pr(>F)"] %>% format.pval(pv = ., eps = .001, digits = 2)`). 

```{r}
mod %>% anova()
```

## Mean comparisons

Following a significant F-test, one will want to compare genotype means.

```{r}
mean_comparisons <- mod %>% 
  emmeans(pairwise ~ "variety", adjust="tukey") %>% 
  pluck("emmeans") %>% 
  cld(details=TRUE, Letters=letters) # add letter display

# If cld() does not work, try CLD() instead.
# Add 'adjust="none"' to the emmeans() and cld() statement
# in order to obtain t-test instead of Tukey!

mean_comparisons$emmeans # adjusted variety means

mean_comparisons$comparisons # differences between adjusted variety means 
```

# Present results

## Mean comparisons

For this example we can create a plot that displays both the raw data and the results, *i.e.* the comparisons of the adjusted means that are based on the linear model.

```{r}
ggplot() +
  # black dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = variety)
  ) +
  # red dots representing the adjusted means
  geom_point(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = variety),
    color = "red",
    position = position_nudge(x = 0.1)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = mean_comparisons$emmeans,
    aes(ymin = lower.CL, ymax = upper.CL, x = variety),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.1)
  ) +
  # red letters 
  geom_text(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = variety, label = .group),
    color = "red",
    position = position_nudge(x = 0.2)
  ) + 
  ylim(0, NA) + # force y-axis to start at 0
  ylab("Yield in t/ha") + # label y-axis
  xlab("Variety") +      # label x-axis
  labs(caption = "Black dots represent raw data
       Red dots and error bars represent adjusted mean with 95% confidence limits per variety
       Means followed by a common letter are not significantly different according to the Tukey-test") +
  theme_classic() # clearer plot format 
```

# Exercises

## Exercise 1

This example is taken from *"Example 5.4"* of the course material ["Quantitative Methods in Biosciences (3402-420)"](https://www.uni-hohenheim.de/en/module-catalogue/modul/quantitative-methods-in-biosciences-3402-420){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Mead et al. (1993, p.54)](https://www.routledge.com/Statistical-Methods-in-Agriculture-and-Experimental-Biology/Mead/p/book/9781584881872){target="_blank"}. The percentage moisture content is determined from multiple samples for each of four different soils. Notice that for this dataset you have no information on any specific field trial layout (*i.e.* `row` or `col` column are not present in the dataset). Therefore you should skip trying to create a field layout with `desplot()` and instead focus on the following:

* Explore
  + How many samples per soil were taken?
  + Which soil has the highest value for moisture?
  + Draw a plot with moisture values per soil
* Analyze
  + Compute an ANOVA 
  + Perform multiple (mean) comparisons using the LSD test/t-test.
* Repeat the analysis, but this time remove all moisture values larger than 12 at the very beginning.

```{r}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Mead1993b.csv"
ex1dat <- read_csv(dataURL)
```

**R-Code and exercise solutions**

Please [click here](https://github.com/SchmidtPaul/DSFAIR/tree/master/Rpurl){target="_blank"} to find a folder with `.R` files. Each file contains

 * the entire R-code of each example combined, including
 * solutions to the respective exercise(s).

```{r, echo=FALSE, eval=FALSE}
# exercise 1

# packages
pacman::p_load(readr, tidyverse, forcats, # data import and handling
               emmeans, multcomp, multcompView, # mean comparisons
               ggplot2, desplot,          # plots
               report, equatiomatic)      # automated analysis summaries

# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Mead1993b.csv"
ex1dat <- read_csv(dataURL)

# in case of removing all moisture values larger than 12 run this:
# ex1dat <- ex1dat %>% 
#   filter(moisture <= 12)

# format first!
ex1dat <- ex1dat %>% 
  mutate_at(vars(soil), as.factor)

# How many samples per soil were taken?
ex1dat %>% 
  count(soil)

# Which soil has the highest value for moisture?
ex1dat %>% 
  group_by(soil) %>% 
  summarise(mean = mean(moisture)) %>% 
  arrange(desc(mean))

# Draw a plot with moisture values per soil
ggplot(data = ex1dat,
       aes(y = moisture, x = soil)) +
  geom_point() +
  ylim(0, NA) +
  theme_classic()

# Compute an ANOVA
mod <- lm(moisture ~ soil, data = ex1dat)

mod %>% anova() # significant!

# Perform multiple (mean) comparisons using the LSD test/t-test.
mod %>% 
  emmeans(pairwise ~ "soil", adjust="none") %>% # adjust="none" for t-test
  pluck("emmeans") %>% 
  cld(details=TRUE, Letters=letters)
```

