---
title: "Latin square design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(readr, tidyverse,     # data import and handling
               emmeans, multcomp,    # mean comparisons
               ggplot2, desplot,     # plots
               report, equatiomatic) # automated analysis summaries
```

# Data

This example is taken from the `agridat`*: Agricultural Datasets* package. It considers data published in [Bridges (1989)](https://rdrr.io/github/kwstat/agridat/man/bridges.cucumber.html){target="_blank"} from a cucumber yield trial set up as a latin square design. Notice that the original dataset considers two trials (at two locations), but we will focus on only a single trial here.  

## Import

```{r, message=FALSE, warning=FALSE}
# data (from agridat package)
dat <- agridat::bridges.cucumber %>% 
  filter(loc == "Clemson") %>% # subset data from only one location
  dplyr::select(-loc) # remove loc column which is now unnecessary

dat
```

## Formatting

While `gen` is already correctly encoded as factor, the columns `row` and `col` should be encoded as factors, too, since R by default encoded them as integer. However, we also want to keep `row` and `col` as integer for the `desplot()` function. Therefore we create copies of these columns encoded as factors and named `rowF` and `colF`

```{r}
dat <- dat %>% 
  as_tibble() %>% # tibble data format for convenience
  mutate(rowF = row %>% as.factor,
         colF = col %>% as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `col` of each plot in the trial. 

```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
desplot(data = dat,
        form = gen ~ row + col, flip = TRUE,
        out1 = row, out1.gpar=list(col="black", lwd=3),
        out2 = col, out2.gpar=list(col="black", lwd=3),
        text = gen, cex = 1, shorten = "no",
        main = "Field layout", 
        show.key = FALSE)
```

We could also have a look at the arithmetic means and standard deviations for yield per `genotype`, `row` or `col`.

```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(gen) %>% # genotype
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(desc(mean))
```

<div class = "row"> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(row) %>% # row
  summarize(mean    = mean(yield),
            std.dev = sd(yield))
```
</div> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(col) %>% # column
  summarize(mean    = mean(yield),
            std.dev = sd(yield))
```
</div> </div>

We can also create a plot to get a better feeling for the data.

```{r, fig.height = 3, fig.width = 4, fig.align = "center"}
ggplot(data = dat,
       aes(y = yield, x = gen, color = rowF, shape=colF)) +
  geom_point(size=2) +  # scatter plot with larger points
  ylim(0, NA) +   # force y-axis to start at 0
  theme_classic() # clearer plot format 
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable and (fixed) `gen` effects, as well as `rowF` and `colF` effects. **Important:** Don't forget to use the variables for rows and columns that are encoded as factors and thus not the ones used in the `desplot()` function above.

```{r}
mod <- lm(yield ~ gen + rowF + colF, data = dat)
```

## ANOVA

Moreover, we can conduct an ANOVA for this model.

```{r}
mod %>% anova()
```

As can be seen, the F-test finds the `gen` effects to be statistically significant (p = `r anova(mod)["gen", "Pr(>F)"] %>% round(3)` < 0.05). 

## Mean comparisons

Following a significant F-test, one will want to compare genotype means.

```{r}
mean_comparisons <- mod %>% 
  emmeans(pairwise ~ "gen", adjust="tukey") %>% # adjust="none" for t-test
  pluck("emmeans") %>% 
  cld(details=TRUE, Letters=letters) # add letter display

mean_comparisons$emmeans # adjusted genotype means

mean_comparisons$comparisons # differences between adjusted genotype means 
```

# Present results

## Mean comparisons

For this example we can create a plot that displays both the raw data and the results, *i.e.* the comparisons of the adjusted means that are based on the linear model.

```{r}
ggplot() +
  # black dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = gen)
  ) +
  # red dots representing the adjusted means
  geom_point(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = gen),
    color = "red",
    position = position_nudge(x = 0.1)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = mean_comparisons$emmeans,
    aes(ymin = lower.CL, ymax = upper.CL, x = gen),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.1)
  ) +
  # red letters 
  geom_text(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = gen, label = .group),
    color = "red",
    position = position_nudge(x = 0.2)
  ) + 
  ylim(0, NA) + # force y-axis to start at 0
  ylab("Yield") + # label y-axis
  xlab("Cucumber genotype") + # label x-axis
  labs(caption = "Black dots represent raw data
       Red dots and error bars represent adjusted mean with 95% confidence limits per genotype
       Means followed by a common letter are not significantly different according to the Tukey-test") +
  theme_classic() # clearer plot format 
```

## dataset

```{r results='asis'}
dat %>% 
  dplyr::select(-row, -col) %>% 
  report() %>% text_short()
```

## anova

```{r results='asis'}
mod %>% 
  anova %>% 
  report() %>% text_short()
```