---
title: "RCBD with 2 factors"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r}
# packages
pacman::p_load(tidyverse, # data import and handling
               conflicted, # handling function conflicts
               emmeans, multcomp, multcompView, # adjusted mean comparisons
               ggplot2, desplot) # plots
```

# Data

This data is a slightly modified version of that in [the split-plot chapter](splitplot_GomezGomez1984.html) published in [Gomez & Gomez (1984)](https://www.wiley.com/en-us/Statistical+Procedures+for+Agricultural+Research%2C+2nd+Edition-p-9780471870920) from a yield (kg/ha) trial with 4 genotypes (`G`) and 6 nitrogen levels (`N`), leading to 24 treatment level combinations. The data set here has 3 complete replicates (`rep`) and is laid out as a randomized complete block design (RCBD).

## Import
```{r}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/riceRCBD.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting

Before anything, the columns `rep`, `N` and `G` should be encoded as factors, since R by default encoded them as character.

```{r}
dat <- dat %>% 
  mutate_at(vars(rep, N, G), as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `col` of each plot in the trial. 

```{r, fig.height = 3, fig.align = "center"}
desplot(data = dat,
        form = rep ~ col + row | rep, # fill color per rep, headers per rep
        text = G, cex = 1, shorten = "no", # show genotype names per plot
        col  = N, # color of genotype names for each N-level
        out1 = col, out1.gpar = list(col = "darkgrey"), # lines between columns
        out2 = row, out2.gpar = list(col = "darkgrey"), # lines between rows
        main = "Field layout", show.key = TRUE, key.cex = 0.7) # formatting
```

Just as in a RCBD for a single treatment factor, a RCBD for two treatment factors has replicates arranged as complete blocks. Thus, all 24 treatment level combinations are grouped together in three complete blocks, respectively. 

We could now have a look at the arithmetic means and standard deviations for yield per genotype (`G`) and nitrogen level (`N`) separately, but also for their combinations:

```{r}
dat %>% 
  group_by(G) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(desc(mean))

dat %>% 
  group_by(N) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(desc(mean))

dat %>% 
  group_by(N, G) %>% 
  summarize(mean    = mean(yield),
            std.dev = sd(yield)) %>% 
  arrange(desc(mean)) %>% 
  print(n=Inf) # show more than default 10 rows
```

We can also create a plot to get a better feeling for the data.

```{r, fig.align = "center"}
ggplot(
  data = dat,
  aes(
    y = yield,
    x = N,
    color = N
  )
) +
  facet_wrap(~G, labeller = label_both) + # facette per G level
  geom_point() + # dots representing the raw data
  scale_y_continuous(
    limits = c(0, NA), # make y-axis start at 0
    expand = expansion(mult = c(0, 0.1)) # no space below 0
  ) +
  scale_x_discrete(name = NULL) + # x-axis
  theme_bw() + # clearer plot format
  theme(legend.position = "bottom") # legend on top
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable. In this example it makes sense to mentally group the effects in our model as either *design effects* or *treatment effects*. The *treatments* here are the genotypes `G` and the nitrogen levels `N` which we will include in the model as main effects, but also via their interaction effect `N:G`. Regarding the *design*, the model needs to contain a block (`rep`) effect.

```{r}
mod <- lm(
  yield ~ N + G + N:G + rep,
  data = dat
)
```

<aside>
Hint: `N*G` is short for `N + G + N:G`.
</aside>

## ANOVA

As a next step, we can conduct an ANOVA for this model. 

```{r}
anova(mod)
```

Whenever we have multiple treatment effects in a model, we should focus on the term with the highest complexity when it comes to the ANOVA. In this case, this is the `N:G` interaction effect and its F-test is significant (p<0.001). 

As a result, we should only compares means for the 24 genotype-nitrogen-combinations ("cell means") and not for the 6 nitrogens across the genotypes or 4 genotypes across the nitrogens ("marginal means"). This is because marginal means are misleading in the presence of interaction. 

<aside>
Do not confuse "marginal means" with "estimated marginal means (emmeans)". Both, cell means and marginal means can be estimated via `emmeans::emmeans()`.
</aside>

## Mean comparisons

Accordingly, we will proceed comparing the 24 means for `N:G`. Whenever means of a two-way treatment interaction such as this are to be compared, I choose between two options, which are both valid but one may be more apt to answer the question at hand:

 1. `~ N:G` Comparing all combinations to all other combinations
 2. `~ N|G` Comparing all genotype means per nitrogen (or the other way around)
 
<aside>
See also [this answer on stackoverflow](https://stackoverflow.com/a/70863531/8830099).
</aside>

### Option 1: `~ N:G`

Here, we really calculate and test all of the 276 possible differences between the 24 nitrogen-genotype combination means.

```{r}
all_mean_comparisons <- mod %>%
  emmeans(specs = ~ N:G) %>% # compare all combs to all other combs
  cld(adjust = "none", Letters = letters) # add compact letter display

all_mean_comparisons
```


```{r, fig.align = "center", class.source = 'fold-hide'}
all_mean_comparisons <- all_mean_comparisons %>% 
  as_tibble() %>%
  mutate(N_G = paste0(N, "-", G)) %>% # create helper column with combs
  mutate(N_G = fct_reorder(N_G, emmean)) # sort combs according to emmean

# do the same for the raw data
dat <- dat %>% 
  mutate(N_G = paste0(N, "-", G)) %>% 
  mutate(N_G = fct_relevel(N_G, levels(all_mean_comparisons$N_G)))


ggplot() +
  # dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = N_G, color = N),
    position = position_nudge(x = -0.2)
  ) +
  # black boxplot
  geom_boxplot(
    data = dat,
    aes(y = yield, x = N_G),
    width = 0.05,
    outlier.shape = NA,
    position = position_nudge(x = -0.1)
  ) +
  # red mean value
  geom_point(
    data = all_mean_comparisons,
    aes(y = emmean, x = N_G),
    size = 2,
    color = "red"
  ) +
  # red mean errorbar
  geom_errorbar(
    data = all_mean_comparisons,
    aes(ymin = lower.CL, ymax = upper.CL, x = N_G),
    width = 0.05,
    color = "red"
  ) +
  # red letters
  geom_text(
    data = all_mean_comparisons,
    aes(
      y = 10000,
      x = N_G,
      label = .group
    ),
    angle = 90,
    hjust = 0,
    color = "red"
  ) +
  # y-axis
  scale_y_continuous(
    name = "Yield",
    limits = c(0, 12500),
    expand = expansion(mult = c(0, 0.1))
  ) +
  # x-axis
  scale_x_discrete(name = "Nitrogen-Genotype combination") +
  # general layout
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    legend.position = "bottom"
  ) +
  labs(
    caption = str_wrap("Black dots represent raw data. Red dots and error bars represent estimated marginal means ± 95% confidence interval per group. Means not sharing any letter are significantly different by the t-test at the 5% level of significance.", width = 120)
  )
```

As can be seen, this option can lead to overwhelming results when there are too many treatment level combinations.

### Option 2: `~ N|G`

Here, we only calculate and test 60 differences between the 24 nitrogen-genotype combination means, since - separately for each of the 4 genotypes - we only compare all 6 nitrogen means to each other, leading to 4x15=60 comparisons. (Note that you may analogously switch `N|G` to `G|N` and instead present results for these 6x6=36 comparisons, if they seem more apt for your research question.)

```{r}
withinG_mean_comparisons <- mod %>%
  emmeans(specs = ~ N|G) %>% 
  cld(adjust="none", Letters=letters) # add compact letter display

withinG_mean_comparisons 
```

```{r, fig.align = "center", class.source = 'fold-hide'}
withinG_mean_comparisons <- as_tibble(withinG_mean_comparisons)

ggplot() +
  facet_wrap(~G, labeller = label_both) + # facette per G level
  # dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = N, color = N)
  ) +
  # red dots representing the adjusted means
  geom_point(
    data = withinG_mean_comparisons,
    aes(y = emmean, x = N),
    color = "red",
    position = position_nudge(x = 0.1)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = withinG_mean_comparisons,
    aes(ymin = lower.CL, ymax = upper.CL, x = N),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.1)
  ) +
  # red letters
  geom_text(
    data = withinG_mean_comparisons,
    aes(y = emmean, x = N, label = str_trim(.group)),
    color = "red",
    hjust = 0,
    position = position_nudge(x = 0.2)
  ) +
  # y-axis
  scale_y_continuous(
    name = "Yield",
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  # x-axis
  scale_x_discrete(name = NULL) +
  # general layout
  theme_bw() + # clearer plot format
  theme(legend.position = "bottom") + # legend on top
  labs(caption = str_wrap("The four facettes represent genotypes A, B, C and D. Black dots represent raw data. Red dots and error bars represent estimated marginal means ± 95% confidence interval per group. For each genotype separately, means not sharing any letter are significantly different by the t-test at the 5% level of significance.", width = 120))
```

**R-Code and exercise solutions**

Please [click here](https://github.com/SchmidtPaul/DSFAIR/tree/master/Rpurl){target="_blank"} to find a folder with `.R` files. Each file contains

 * the entire R-code of each example combined, including
 * solutions to the respective exercise(s).