---
title: "Row-column design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(janitor, readr, tidyverse, # data import and handling
               lme4,                      # linear mixed model 
               emmeans, multcomp,         # mean comparisons
               ggplot2, desplot)          # plots 
```

# Data

This example is taken from Chapter *"3.10 Analysis of a resolvable row-column design"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Kempton and Fox (1997)](https://www.springer.com/de/book/9780412547508){target="_blank"} from a yield trial laid out as a resolvable row-column design. When two blocking structures are used which are arranged in rows and columns, we have a *row-column design*. If rows and columns can be grouped to form complete replicates, the design is *resolvable*, as in case of one-way blocking. The trial had 35 genotypes (`gen`), 2 complete replicates (`rep`) with 5 rows (`row`) and 7 columns (`col`). Thus, a complete replicate is subdivided into incomplete rows and columns.

