---
title: "Row-column design"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html		
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(purrr)
```

```{r, message=FALSE, warning=FALSE}
# packages
pacman::p_load(janitor, readr, tidyverse, # data import and handling
               lme4, lmerTest,            # linear mixed model 
               emmeans, multcomp,         # mean comparisons
               ggplot2, desplot)          # plots 
```

# Data

This example is taken from Chapter *"3.10 Analysis of a resolvable row-column design"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data published in [Kempton and Fox (1997)](https://www.springer.com/de/book/9780412547508){target="_blank"} from a yield trial laid out as a resolvable row-column design. The trial had 35 genotypes (`gen`), 2 complete replicates (`rep`) with 5 rows (`row`) and 7 columns (`col`). Thus, a complete replicate is subdivided into incomplete rows and columns.

## Import

```{r, message=FALSE, warning=FALSE}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Kempton%26Fox1997.csv"
dat <- read_csv(dataURL)

dat
```

## Formatting
Before anything, the columns `gen` and `rep` should be encoded as factors, since R by default encoded them as character.

```{r}
dat <- dat %>% 
  mutate_at(vars(gen, rep), as.factor)
```

## Exploring

In order to obtain a field layout of the trial, we can use the `desplot()` function. Notice that for this we need two data columns that identify the `row` and `col` of each plot in the trial. 

```{r, fig.height = 3, fig.align = "center"}
desplot(data = dat,
        form = gen ~ col + row | rep,            # fill color per genotype, headers per replicate
        text = gen, cex = 0.7, shorten = "no",   # show genotype names per plot
        out1 = row, out1.gpar=list(col="black"), # lines between rows
        out2 = col, out2.gpar=list(col="black"), # lines between columns
        main = "Field layout", show.key = F)     # formatting
```

When two blocking structures are used which are arranged in rows and columns, we have a *row-column design*. If rows and columns can be grouped to form complete replicates, the design is *resolvable*, as in case of one-way blocking.

We could also have a look at the arithmetic means and standard deviations for yield per genotype (`gen`).

```{r, warning=FALSE, message=FALSE}
dat %>% 
  group_by(gen) %>% 
  summarize(mean    = mean(yield, na.rm=TRUE),
            std.dev = sd(yield, na.rm=TRUE)) %>% 
  arrange(desc(mean)) %>% # sort
  print(n=Inf) # print full table
```

We can also create a plot to get a better feeling for the data.

```{r, fig.align = "center", warning=FALSE, message=FALSE}
ggplot(data = dat,
       aes(y = yield, x = gen, color=rep)) +
  geom_point() +  # scatter plot
  ylim(0, NA) +   # force y-axis to start at 0
  theme_classic() + # clearer plot format 
  theme(axis.text.x = element_text(angle=90, vjust=0.5), # rotate x-axis label
        panel.grid.major.x = element_line(), # add vertikal grid lines
        legend.position = "top") # legend on top 
```

# Modelling

Finally, we can decide to fit a linear model with `yield` as the response variable and fixed `gen` and `rep` effects. There also need to be terms for the 5 rows (`row`) and 7 columns (`col`) per replicate. Notice that they can either be taken as a fixed or a random effects. Since our goal is to compare genotypes, we will determine which of the two models we prefer by comparing the average standard error of a difference (s.e.d.) for the comparisons between adjusted genotype means - the lower the s.e.d. the better.

Also notice that in our dataset the `row` and `col` columns were defined as *numeric* because this format is needed when creating the field layout with desplot as we did above. However, now in order to have row and column effects in our model, we need to define them as factors. To make this clear, we will not just change the format, but create a copy of these two columns in the correct format:

```{r}
dat <- dat %>% 
  mutate(row_fct = as.factor(row),
         col_fct = as.factor(col))
```

Furthermore, when building our model we must ensure that row and column effects are being estimated **per replicate**. Thuse, we have 2x5 row effects and 2x7 column effects. This can be done as:

<div class = "row"> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
# rows and cols fixed (linear model)
mod.fb <- lm(yield ~ gen + rep +
               rep:row_fct + 
               rep:col_fct,
             data = dat)

mod.fb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pluck("SE") %>% # extract s.e.d. column
  mean() # get arithmetic mean
```
</div> <div class = "col-md-6">
```{r, warning=FALSE, message=FALSE}
# rows and cols random (linear mixed model)
mod.rb <- lmer(yield ~ gen + rep +
                 (1|rep:row_fct) + 
                 (1|rep:col_fct),
               data = dat)

mod.rb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("contrasts") %>% # extract diffs
  as_tibble %>% # format to table
  pluck("SE") %>% # extract s.e.d. column
  mean() # get arithmetic mean
```
</div> </div>

As a result, we find that the model with random row and column effects has the smaller s.e.d. and is therefore more precise in terms of comparing genotypes.

## ANOVA

Thus, we can conduct an ANOVA for this model. As can be seen, the F-test of the ANOVA (using Kenward-Roger’s method for denominator degrees-of-freedom and F-statistic) finds the `gen` effects to be statistically significant (p = `r anova(mod.rb, ddf="Kenward-Roger")["gen", "Pr(>F)"] %>% format.pval(pv = ., eps = .001, digits = 2)` < 0.05). 

```{r}
mod.rb %>% anova(ddf="Kenward-Roger")
```

## Mean comparisons

```{r}
mean_comparisons <- mod.rb %>%
  emmeans(pairwise ~ "gen",
          adjust = "tukey",
          lmer.df = "kenward-roger") %>%
  pluck("emmeans") %>%
  cld(details = TRUE, Letters = letters) # add letter display

mean_comparisons$emmeans # adjusted genotype means
```

# Present results

## Mean comparisons

For this example we can create a plot that displays both the raw data and the results, *i.e.* the comparisons of the adjusted means that are based on the linear model.

```{r}
# sort gen factors according to emmean
# for adjusted means
mean_comparisons$emmeans <- mean_comparisons$emmeans %>% 
  mutate(gen = fct_reorder(.f = gen, .x = emmean))

# for raw data
dat <- dat %>% 
  mutate(gen = fct_relevel(.f = gen, 
                           as.character(mean_comparisons$emmeans$gen)))

ggplot() +
  # black dots representing the raw data
  geom_point(
    data = dat,
    aes(y = yield, x = gen)
  ) +
  # red dots representing the adjusted means
  geom_point(
    data = mean_comparisons$emmeans,
    aes(y = emmean, x = gen),
    color = "red",
    position = position_nudge(x = 0.1)
  ) +
  # red error bars representing the confidence limits of the adjusted means
  geom_errorbar(
    data = mean_comparisons$emmeans,
    aes(ymin = lower.CL, ymax = upper.CL, x = gen),
    color = "red",
    width = 0.1,
    position = position_nudge(x = 0.1)
  ) +
  # red letters 
  geom_text(
    data = mean_comparisons$emmeans,
    aes(y = lower.CL, x = gen, label = .group),
    color = "red",
    angle = 90,
    hjust = 1,
    position = position_nudge(y = - 0.1)
  ) + 
  ylim(0, NA) + # force y-axis to start at 0
  ylab("Yield in t/ha") + # label y-axis
  xlab("Genotype") +      # label x-axis
  labs(caption = "Black dots represent raw data
       Red dots and error bars represent adjusted mean with 95% confidence limits per genotype
       Means followed by a common letter are not significantly different according to the Tukey-test") +
  theme_classic() + # clearer plot format 
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) # rotate x-axis label
```

# Exercises

## Exercise 1
This example is taken from *"Example 8"* of the course material ["Mixed models for metric data (3402-451)"](https://www.uni-hohenheim.de/en/module-catalogue/lehrveranstaltung/mixed-models-for-metric-data-3402-451){target="_blank"} by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257){target="_blank"}. It considers data from a yield trial with oat. There were 64 genotypes laid out as a 8 x 8 lattice with 3 replicates.

* Explore
  + Which genotypes have missing yield observations?
  + Create a field trial layout using ´desplot()´ where the plots are filled according to their yield. Try to see whether some areas have lower/higher yields.  
* Analyze
  + Set up the model with random incomplete block effects.
  + Extract the variance components.
  + Compute an ANOVA 
  + Perform multiple (mean) comparisons using the LSD test/t-test.

```{r, warning=FALSE, message=FALSE}
# data (import via URL)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/RowColFromUtz.csv"
ex1dat <- read_csv(dataURL)
```

